---
title: "Covid_vaccination_rates"
author: "Kelly_F"
date: "12/3/2021"
output:
  github_document: default

---
# Import Vaccination Data 

Import CA data on COVID vaccines administered by zipcode from: https://data.ca.gov/dataset/covid-19-vaccine-progress-dashboard-data-by-zip-code/resource/15702a90-aa5d-49bc-8621-a8129630725a 

NOTE: answers in this lab report may differ from answer key on lab walk through b/c data has been updated since last week as of 11/30/21. 

```{r}
# Import CSV
vax <- read.csv("./covid19vaccinesbyzipcode_test.csv", header = TRUE)
head(vax)
```
> Q1. What column details the total number of people fully vaccinated? 

```{r}
head(vax$persons_fully_vaccinated)
```
> Q2. What column details the Zip code tabulation area? 

```{r}
head(vax$zip_code_tabulation_area)
```
> Q3. What is the earliest date in this dataset? 
"2021-01-05"

```{r}
head(vax$as_of_date)
```
> Q4. What is the latest date in this dataset?
"2021-11-30". Note, this differs from lab report solutions because the dataset has been updated since 11/24 when class was held. 

```{r}
dim(vax)
vax$as_of_date[84672]
```

# Look at overview of df

```{r}
# Install and load Skimr package
#install.packages("skimr")
library(skimr)

#skimr::skim(vax)
```

> Q5. How many numeric columns are in this dataset? 
9 

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 
8472. This numer was taken from the skimr report, from the "n_missing" column. Note, this differs from lab report solutions because the dataset has been updated since 11/24 when class was held. 

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 
10.01 %

```{r}
# Find dimension of data frame
dim(vax)

# Divide # of missing values for persons_fully_vaccinated by number of columns
round(((8472/84672)*100), 2)
```
# Working with Dates

We need to change dates in our vax$as_of_date column to useful format using the lubridate package so that we can work with our data in R

```{r}
library(lubridate)

# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

> How many days have passed since the first vaccination reported in this dataset?

```{r}
# Now that dates have been reformatted, we can now do math with our as_of_date column 

today() - vax$as_of_date[1]
```
> how many days does the the dataset span?

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```
> Q9. How many days have passed since the last update of the dataset?
3 days

```{r}
today()- vax$as_of_date[nrow(vax)]
```
> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?
48

```{r}
length(unique(vax$as_of_date))
```

# Working with Zip Codes 

Will be using zipcodeR package to make working with zip codes easier

```{r}
# install.packages("zipcodeR")
library(zipcodeR)
```
Note: reverse_zipcode function, which gathers census data, can be used to investigate if various soceioeconomic factors correlate with vaccination rates

```{r}
# Calculate the distance between the centroids of two zip codes. E.g.: 
zip_distance('92037','92109')
```

```{r}
# Pull census data for the following zip codes
reverse_zipcode(c('92037', "92109") )
```

# Focus on the San Diego Area

```{r}
# Subset to San Diego county only areas using base R
sd <- vax[ vax$county == "San Diego", ]

# Alternatively, we could have subset to San Diego using dplyr: 
library(dplyr)
#sd <- filter(vax, county == "San Diego")

nrow(sd)

```
> Q11. How many distinct zip codes are listed for San Diego County? 
107

```{r}
length(unique(sd$zip_code_tabulation_area))
```
> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?
92154

```{r}
sd$zip_code_tabulation_area[which.max(sd$age12_plus_population)]
```
Using dplyr select all San Diego “county” entries on “as_of_date” “2021-11-09” and use this for the following questions.

```{r}
# Data on as of date nov 9th
sd_nov_9 <- filter(sd, as_of_date == "2021-11-09")
```
 > Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2021-11-09”?
 67.41%
 
```{r}
round((mean(sd_nov_9$percent_of_population_fully_vaccinated, na.rm=TRUE)*100), 2)
```
> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2021-11-09”?

```{r}
hist(sd_nov_9$percent_of_population_fully_vaccinated, 
     main="Histogram of COVID Vaccine Rates in San Diego County",
        xlab="% of population fully vaccinated on Nov 9th, 2021")
```

# Focus on UCSD/La Jolla 

UCSD area code is 92037

```{r}
# Filter to UCSD data
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
```

> Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
library(ggplot2)

# Graph time course of vaccination rate for UCSD
ggplot(ucsd) +
  aes(x= as_of_date, y= percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated", 
       title= "Vaccination Time Course for UCSD/La Jolla Zip Code 92037")
```
# Comparing La Jolla/UCSD to similar sized areas

Compare to area codes with similarly sized population. 

Go back to the original dataset and compare to area codes that have a population at least as large as La Jolla
```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2021-11-16")

head(vax.36)

```
> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
mean_pop <- mean(vax.36$percent_of_population_fully_vaccinated)

# Add mean_pop to ggplot graph of vaccination timecourse for UCSD/LA Jolla county

ggplot(ucsd) +
  aes(x= as_of_date, y= percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  geom_hline(yintercept = mean_pop, color="red")+
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated", 
       title= "Vaccination Time Course for UCSD/La Jolla Zip Code 92037")
```
> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```
> Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36,
       aes(percent_of_population_fully_vaccinated))+ 
  geom_histogram(bins=20)
```
> Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?
The average % of population fully vaccinated for area codes w/ a population at least as large as san diego on  11/16/21 is 66.45%. Based on this, 92040 area code has a lower than average vaccination rate (~52%), and 92109 has an above average vaccination rate (~69%). 

```{r}
# % fully vaccinated for 92040 on 11/16/21
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)

# % fully vaccinated for 92109 on 11/16/21
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)

```
> Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
# Subset data to populations larger than 36K
vax.36.all <- filter(vax, age5_plus_population > 36144)

# Plot vaccine timecourse for all area codes. 
ggplot(vax.36.all) +
  aes(x=as_of_date,
      y=percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(0,1) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination Rates Across CA",
       subtitle="Only zip codes w/ population above 36K are shown") +
  geom_hline(yintercept = mean_pop, linetype=2)
```
> Q21. How do you feel about traveling for Thanksgiving and meeting for in-person class next Week?
About as good as I can given the circumstances... :D
