---
title: "lab_6_supplement Writing Functions"
author: "Kelly_F"
date: "10/20/2021"
output: html_document
---
**Section 1:** Improving analysis code by writing functions
A. Improve this regular R code by abstracting the main activities in your own new function. Note, we will go through this example together in the formal lecture. The main steps should entail running through the code to see if it works, simplifying to a core working code snippet, reducing any calculation duplication, and finally transferring your new streamlined code into a more useful function for you.

```{r}
# (A. Can you improve this analysis code?
df <- data.frame(a=1:10, b=seq(200,400,length=10),c=11:20,d=NA) #create df
df$a <- (df$a - min(df$a)) / (max(df$a) - min(df$a))
df$b <- (df$b - min(df$a)) / (max(df$b) - min(df$b))
df$c <- (df$c - min(df$c)) / (max(df$c) - min(df$c))
df$d <- (df$d - min(df$d)) / (max(df$a) - min(df$d))

# Assuming that the code is meant to use the same input $column for all min & max functions per line above, and that typos exist, I would correct the code to the following 
mask <- data.frame(a=1:10, b=seq(200,400,length=10),c=11:20,d=NA) =

# Write the function. Function makes all columns of dataframe the same. 
df_analysis <- function(column){
  column <- (column - min(column)) / (max(column) - min(column))
}

# Test function on one column of data frame 
print(df_analysis(mask$a))

# Apply function to entire data frame
analyzed_mask <- apply(mask, 2, df_analysis)
analyzed_mask

```
B. Next improve the below example code for the analysis of protein drug interactions by abstracting the main activities in your own new function. Then answer questions 1 to 6 below. It is recommended that you start a new Project in RStudio in a new directory and then install the bio3d package noted in the R code below (N.B. you can use the command install.packages("bio3d") or the RStudio interface to do this).

Then run through the code to see if it works, fix any copy/paste errors before simplifying to a core working code snippet, reducing any calculation duplication, and finally transferring it into a more useful function for you.

**Q1. What type of object is returned from the read.pdb() function?** 
List 

**Q2. What does the trim.pdb() function do?**
The trim.pdb() function produces a new PDB object that contains a subset of atoms from a larger PDB object.

**Q3. What input parameter would turn off the marginal black and grey rectangles in the plots and what do they represent in this case?**
bot=FALSE. They represent the residues of secondary structure object, which in this case is ____? 

**Q4. What would be a better plot to compare across the different proteins?**

**Q5. Which proteins are more similar to each other in their B-factor trends. How could you quantify this?**

**Q6. How would you generalize the original code above to work with any set of input
protein structures?**

```{r}
# install.packages("bio3d")

# Can you improve this analysis code?
library(bio3d)
s1 <- read.pdb("4AKE")  # kinase with drug
s2 <- read.pdb("1AKE")  # kinase no drug
s3 <- read.pdb("1E4Y")  # kinase with drug

s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s3, chain="A", elety="CA") #typo fixed, s1 -> s3

s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b

plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")

#1 Type of object returned from read.pdb
typeof(s1)

#2 Determine what trim.pdb function does 
?trim.pdb

# Create function to simplify code from above 
plot_Bfactor <- function(PDB_code){
  protein <-read.pdb(PDB_code) 
  protein.chainA <- trim.pdb(protein, chain="A", elety="CA")
  protein.b <- protein.chainA$atom$b
  plotb3(protein.b, sse=protein.chainA, typ="l", ylab="Bfactor", bot=FALSE) #rectangles at bottom of the graph are removed w/ bot=FALSE
}

# See if function works on single protein
plot_Bfactor("4AKE")

# Create list of all proteins you want to run through function 
proteins <- c("4AKE", "1AKE", "1E4Y")

# Apply function to all proteins of interest 
lapply(proteins, plot_Bfactor)


```

