---
title: "10_13_21_writing_R_functions"
author: "Kelly_F"
date: "10/15/2021"
output:
  pdf_document: default
  html_document: default
---

## Quick R Markdown intro

We can write text to be **bolded text**, or *italics*. 

Do (my bullet list example): 

- this
- and that
- and another thing

Note, two lines between text are needed to create a new line. 

--- 

We can include some code: 
```{r}
#option+CMD+I to insert a new code block
#My comment that will not be passed into R

#My Plot
plot(1:10)
```

---
**Q1.** Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an NA value) this can be used as a score to be potentially dropped. Your final function should be adquately explained with code comments and be able to work on an example class gradebook such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]

```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)

sum(is.na(student3))
which(is.na(student3))
mean(na.omit[student3])
mean(student3, na.rm=TRUE)
mean(student3)
# Finding the lowest score of student 1
```


```{r}
min(student1, na.rm=TRUE)

# Find the position of the minimum score
which.min(student1)

# Remove lowest score from student1
student1 <- student1[-which.min(student1)]
student1

# Now take the mean
student1_avg <- mean(student1)
student1_avg

# Repeat example but for student 2. Must remove NA values to calculate mean
# Remove minimum score then calculate mean
student2_avg <- mean(student2[-which.min(student2)],
                     na.rm=TRUE)
student2_avg

# Assign zeros to NA values instead of dropping so that grades aren't inflated. Can then treat zero as a low score to be dropped prior to calculating average
student3[is.na(student3)] <- 0
student3
which.min(student3)
student3_avg <- mean(student3[-which.min(student3)])
student3_avg 

```
```{r}

# Now write a function to calculate overall homework grade 

grade <- function(student_grades){ # A function titled "grade" that needs a vector of student grade scores as an input
  student_grades[is.na(student_grades)] <- 0 # Replace NA values with zero
  overall_grade <- mean(student_grades[-which.min(student_grades)]) # Drop lowest score and calculate the mean
}

# Let's test our function 
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
student2_grade <- grade(student2)
student2_grade
student3_grade <- grade(student3)
student3_grade
```

```{r}
# Now lets apply our grade function this to a df of grades 

# Import data
df <- read.csv("/Volumes/GoogleDrive/My Drive/Courses/BGGN_213/10_15_21_writing_functions/student_homework.csv", header=TRUE, row.names=1, sep=",")

#Test out "grade" function on one row of data 
# Take first row for student1 grades
student1_grades <- as.numeric(df[1,])
student1_grades

# Apply grade function to student1 to calculate overall grade 
student1_overall_grade <- grade(student1_grades)
student1_overall_grade

#Calculate overall grade for all students in the df
for (i in 1:nrow(df)){
  student_grades <- as.numeric(df[i,])
  df$overall.grade[i] <- grade(student_grades)
}
df

# Lets try an alternative approach with the "apply" function 
df2 <- read.csv("/Volumes/GoogleDrive/My Drive/Courses/BGGN_213/10_15_21_writing_functions/student_homework.csv", header=TRUE, row.names=1, sep=",")

overall_grades <- apply(df2, 1, grade)

```
Q2. Using your grade() function and the supplied gradebook, Who is the top scoring student overall in the gradebook? [3pts]

Answer: Student #18

```{r}
# Determine which student the maximum grade comes from
which.max(overall_grades)
```
Q3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained the lowest scores overall? [2pts]

Answer: homework 2 
```{r}
# Replace NA values with zeros in the data frame
mask <- df2
mask[is.na(mask)] <- 0 

# Calculate average score per homework 
avg_homework <- apply(mask, 2, mean)
avg_homework

```
Q4. Optional Extension: From your analysis of the gradebook, which homework was most predictive of overall score (i.e. highest correlation with average grade score)? [1pt]

Answer: homework 5 was the most predictive 

```{r}
# Pearson correlation between homework 1 and overall grades
cor(mask$hw1, overall_grades)

# Use apply function to do all correlations at once
apply(mask, 2, cor, overall_grades)

# Make a graph of our grade scores
boxplot(mask)

```

Write a function "grade2()" to determine an overall grade from a vector of student homework assignment scores OPTIONALLY dropping the lowest single assignment score
```{r}
grade2 <- function(x, drop.lowest=TRUE) {
 x <- as.numeric(x)
 x[ is.na(x)] = 0
 if(drop.lowest) {
 mean(x[ -which.min(x) ])
 } else {
   mean(x)
 }
}
```

